Dynamical modeling of tidal streams

### 2014-01-07 tex/ms.tex
streams: some re-arranging

@@

Because we expect the typical $\Delta\veco$ to scale with the velocity
dispersion of the progenitor, I conjecture that a constant
$\mu_\Omega$ can be used for modeling cold tidal streams of any
(small-ish) velocity dispersion. Further modeling of simulated
streams, however, is necessary to check this and to find the optimal
value of $\mu_\Omega$. Fixing $\mu_\Omega$ leaves a single free
parameter \sigv\ for modeling the frequency offset distribution of a
tidal stream. In modeling to observed data, this parameter needs to be
fit and we expect it to be proportional to the velocity dispersion of
the progenitor, although this proportionality needs to be checked more
carefully (but see Figure 3 in \citealt{Sanders13a}, which shows that
the size of the frequency distribution scales as mass$^{1/3}$, or
approximately as the velocity dispersion). Similarly, the initial
angle distribution, modeled here as an isotropic Gaussian, has a
characteristic spread $\sigma_{\theta}$ which I set here to $0.003$
($=\sigv/[122\kms]$), and we likewise expect this spread to scale with
the velocity dispersion of the progenitor.

@@

@@

\begin{figure*}[t!]
\begin{center}
  \includegraphics[width=0.8\textwidth,clip=]{gd1_evol_lb.eps}
  \includegraphics[width=0.8\textwidth,clip=]{gd1_evol_ld.eps}
  \caption{Simulated stream after $5.011\Gyr$ as a function of
    Galactic longitude. The top panel shows the stream in Galactic
    latitude and the bottom panel shows it in the distance from the
    Sun. The progenitor is shown as the large gray dot and its orbit
    as the dashed lines. The predicted track of the stream is shown as
    the solid line. Insets show various zoomed-in parts of the stream
    to show the stream, progenitor orbit, and predicted stream track
    up close.}\label{fig:gd1_lbd}
%python plot_stream.py ../tex/gd1_evol_lb.ps
%python plot_stream.py ../tex/gd1_evol_ld.ps
\end{center}
\end{figure*}

@@

In our stream model the distribution of $\Delta \veco$ is constant
with time, such that stream members on average are removed to the same
orbit. However, the finite disruption time means that for stream
members to have reached the largest angle offsets from the progenitor,
they have to have been removed onto an orbit that is separated from
the progenitor's by a larger $\Delta \veco$ than average. In our model
this is explicitly expressed by the distribution $p(\Delta
\opar|\Delta \apar)$, calculated in \equationname~(\ref{eq:poparapar})
and shown in the top left panel of
\figurename~\ref{fig:gd1_apar}. Thus, the average orbit changes as a
function of position along the stream, whether expressed as $\Delta
\apar$, $X$, or $l$, and a single-orbit approximation is necessarily
wrong over the full length of the stream.

>>>>

The framework presented in this paper differs from other approaches
that model streams in action--angle coordinates in a few respects. As
in \citet{Helmi99a}, the model is a generative one in that it gives an
analytic prescription for the initial action--angle distribution of
tidal debris that is then evolved in time to generate a tidal
stream. The difference between these two generative models is that I
propose simple, few-parameter models for the initial action--angle
distributions, that I work out the properties of streams in observable
coordinates in more detail, and that I present and use a general
action--angle transformation that allows stream modeling in general
potentials. The average location of a stream in phase-space can be
computed in the framework presented here and it could be used to
replace orbit fitting; \citet{Varghese11a} discussed a similar method
to model observed streams. The method in this paper differs from that
of \citet{Varghese11a} in that the model is fundamentally located in
action--angle space rather than position--velocity space and in that
the width and length of the stream are related to the offsets between
the stream, the progenitor's orbit, and the average orbit as a
function of stream position. Thus, the present framework does not have
to assume a progenitor velocity dispersion, but can constrain it from
the stream data as well. Compared to the recent methods of
\citet{Penarrubia12a}, \citet{Sanders13b}, and \citet{PriceWhelan13a},
which allow for fitting the Milky Way's gravitational potential using
observations of the positions and velocities of individual stream
stars, the method presented here is a generative model and it can be
used with noisy or entirely missing phase-space dimensions.

@@

Because we expect the typical $\Delta\veco$ to scale with the velocity
dispersion of the progenitor, I conjecture that a constant
$\mu_\Omega$ can be used for modeling cold tidal streams of any
(small-ish) velocity dispersion. Further modeling of simulated
streams, however, is necessary to check this and to find the optimal
value of $\mu_\Omega$. In particular, if the progenitor has internal
rotation the action distribution of the debris and $\mu_\Omega$ may be
significantly different. Fixing $\mu_\Omega$ leaves a single free
parameter \sigv\ for modeling the frequency offset distribution of a
tidal stream. In modeling to observed data, this parameter needs to be
fit and we expect it to be proportional to the velocity dispersion of
the progenitor, although this proportionality needs to be checked more
carefully (but see Figure 3 in \citealt{Sanders13a}, which shows that
the size of the frequency distribution scales as mass$^{1/3}$, or
approximately as the velocity dispersion). Similarly, the initial
angle distribution, modeled here as an isotropic Gaussian, has a
characteristic spread $\sigma_{\theta}$ which I set here to $0.003$
($=\sigv/[122\kms]$), and we likewise expect this spread to scale with
the velocity dispersion of the progenitor.

@@

\begin{figure*}[th!!]
\begin{center}
  \includegraphics[width=0.8\textwidth,clip=]{gd1_evol_lb.eps}
  \includegraphics[width=0.8\textwidth,clip=]{gd1_evol_ld.eps}
  \caption{Simulated stream after $5.011\Gyr$ as a function of
    Galactic longitude. The top panel shows the stream in Galactic
    latitude and the bottom panel shows it in the distance from the
    Sun. The progenitor is shown as the large gray dot and its orbit
    as the dashed lines. The predicted track of the stream is shown as
    the solid line. Insets show various zoomed-in parts of the stream
    to show the stream, progenitor orbit, and predicted stream track
    up close.}\label{fig:gd1_lbd}
%python plot_stream.py ../tex/gd1_evol_lb.ps
%python plot_stream.py ../tex/gd1_evol_ld.ps
\end{center}
\end{figure*}

@@

@@

In our stream model the distribution of $\Delta \veco$ is constant
with time, such that stream members on average are removed to the same
orbit. However, the finite disruption time means that for stream
members to have reached the largest angle offsets from the progenitor,
they have to have been removed onto an orbit that is separated from
the progenitor's by a larger $\Delta \veco$ than average. In our model
this is explicitly expressed by the distribution $p(\Delta
\opar|\Delta \apar)$, calculated in \equationname~(\ref{eq:poparapar})
and shown in the top left panel of
\figurename~\ref{fig:gd1_apar}. Thus, the average orbit changes as a
function of position along the stream, whether expressed as $\Delta
\apar$, $X$, or $l$, and a single-orbit approximation is necessarily
wrong over the full length of the stream. This is the case regardless
of the progenitor's orbit.
